% !TEX root = ../thesis.tex
%
\pdfbookmark[0]{Acknowledgement}{Acknowledgement}
\chapter*{Acknowledgement}\label{sec:acknowledgement}
\vspace*{-10mm}

The research reported in this thesis would not have seen daylight without the assistance, patience, and support of many sometimes unwitting individuals, who I feel the urge to thank one by one here.

First off, my ever tireless supervisor, Prof.~Alessio Malizia, who helped me throughout the entire process and pushed me to always nurture my curiosity. Spending time and enjoying an unhealthy amount of coffees with him while listening to his stories was an extremely enriching process that I will always be grateful for.

My second supervisor, Dr.~David Bell, who supported me many times throughout the journey, and --- together with Prof.~Terry Young --- helped me feeling welcome in the Department from the first moment I joined.

Prof.~Alan Dix, who lead me on the far-distant shores of Tiree and deeper within my research and objectives.

Boyce and my office mates in STJN129, as well as my work collegues in WLF\-B\-151, for all the successes, sorrows, and food leftovers we shared over the past years.

Then, on with some more personal notes: to my Mum and Dad, for their continued support at distance and for always being there anytime I needed to get back to recharge my batteries. To Giulia, for standing beside me all along the way and conforting me when I needed the most. To my sister Sara, her husband Samu, and my two precious nieces Matilde and Margherita, for always trying to lit up my mood, no matter what.

And finally, to all my friends at home, with an honourable mention to Alessio, Fabio, Vieri, and Antonio, for making me feel like I never left.

All this won't be forgotten.
